Exploratory data structure comparisons: Three new tools based on Principal Component Analysis

### 2017-03-23 article/article_v0.9.tex
Edited section about PCADSC tools (now section 2)

Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well. However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt. data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable?

Surveys often use mixed modes of administration, e.g. mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results. Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.

The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population. However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a dataset without assuming neither directional nor hierarchical relationships between the variables. We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. Software for using these tools is available in an \texttt{R}-package. \hl{how to refer to our package?}

This manuscript is structured as follows: First, in Section \ref{sec:stateoftheart}, we present the data structure comparison problem in more detail and discuss what statistical methods are already available for solving similar challenges. Next, in Section \ref{sec:pcadscintro}, we move on to a description of the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general. In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries. More specifically, we compare data from Denmark, which has been repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless. At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec:discussion}.

\section{Something about state of the art}
\label{sec:stateoftheart}
\subsection{\hl{More detailed description of the type of problem we wish to address}}
\hl{
\begin{itemize}
\item Two subsets of a dataset, i.e. to datasets with the same variables, but different observations
\item Wish to compare structures without specifying a model or even any variables of interest
\item The most central example is the question of whether the two subsets can readily be combined in a (unknown) data analysis, or if the subset-inducing variable actually implies heterogeneity across the subset division
	\begin{itemize}
		\item Examples: Large scale open source datasets such as the PISA data and ESS (European Social Survey) data and ...(?). In these datasets, the data producers are very far away from the majority of the data analysts. Therefore, problem-specific recommendations about potential instrument-induced challenges in the datasets are not available for the data analysts. How can data producers ensure that this will not be an issue, at least not related to known data gathering differences?
		\item Other examples?
		\item Perhaps a description of what happens if we are to combine the two subsets of the datasets without taking a e.g. an instrument-effect into account. When will it cause problems (maybe: causal graph style?)?
		\item Mention somewhere: We want a solution that is largely independent of the sizes of the two subsets of data. Thereby, a lot of methods that compare each subset to the full dataset in some sense are excluded.
	\end{itemize}
\end{itemize}
}

\subsection{\hl{Describe existing methods used to solve similar questions or parts of the question we are addressing}}
\hl{
\begin{itemize}
\item The simplest case: variable-by-variable tests in distributional differences
	\begin{itemize}
		\item Simple, but scales poorly
		\item Only relates to marginal differences and not to the interplay between variables
	\end{itemize}
\item Karl's papers?
\item Anne's papers: IRT-based methods for surveys
	\begin{itemize}
		\item Moves beyond the marginal approach, but needs a model pre-specified
		\item Thus, it is not a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. This is fundamentally a different (though related) question.
	\end{itemize}
\end{itemize}
}

@@

Though the idea sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once. However, by use of linear algebra, we can decompose and recompose the covariance matrix, such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance. In particular, we find a new representation of the covariance matrix that makes it to gain an overview of the most interesting aspects of the data. On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.

\subsection{Principal component analysis}
Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = \frac{1}{n} \sum_{i=1}^n x_n$ denote their average\hl{(s?)} and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance \hl{Is $\bar{x} \in \RR^d$ with $\bar{x} = (\bar{x_1}, ..., \bar{x_d})^T$?}. Suppose that we want to describe the observations by $q$ numbers instead of the original $d$ numbers. The associated \emph{rank-q-reconstruction error} is defined as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
\emph{Principal component analysis} (PCA) ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, and it provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error. Thus, let $S = U \Lambda U^\top$ be the eigenvalue decomposition of $S$. Here $\Lambda \in \RR^d$ is the diagonal matrix with the eigenvectors $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ in the diagonal, and $U \in \RR^d$ is the orthogonal matrix with the associated eigenvectors $\eta_1,\dotsc,\eta_d \in \RR^d$ in the columns. The eigenvalues are uniquely defined, and the eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different. If some of the eigenvalues are identical, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_j$, then the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation. In practice this only happens if $n < d$, in which case the last $d-n$ eigenvalues will be zero \hl{Anne: What about covariance matrices that are not of full rank (e.g. because of colinearity)? Won't we find eigenvalue multiplicities $> 1$ here as well?}. It is a result from linear algebra that the rank-q-reconstruction error for $q < d$ is achieved for
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
and equals $\sum_{j=q+1}^d \lambda_j$. The eigenvectors $\eta_j \in \RR^d$ are called \emph{loadings}, and the eigenvalues $\lambda_j \ge 0$ may be understood as \emph{variation components}. The projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings are called \emph{scores}. The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$, where the initial subspace is defined as $\hat{K}_0 = \{0\}$, that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
\hl{Anne: Should we maybe assume standardized $x$s here also? I feel like it is part of the PCA algorithm and subtracting the mean in the above seems off for that reason. Also: Something is wrong with the dimensions here. $u$ is $d \times 1$, $x_i$ is $d \times 1$(?) - how do we get $\eta_j$ to also be $d \times 1$?} It is worth emphasizing that the greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension $q$.

%Principal component analysis (PCA) allows us to decompose a data matrix consisting of $n$ observations of $d$ variables into $d$ orthogonal vectors of length $d$. This is in itself not very impressive nor useful. But PCA does not result in just any basis of the data matrix. The basis chosen using PCA has the very favorable property that each vector is chosen such that they maximize the accumulated explained variance. These vectors are usually denoted \textit{principal component scores} or simply \textit{principal components} and they are constructed as linear combinations of the variables in the dataset (or, equivalently, as projections of the data matrix). The coefficients of each variable in these projections are denoted \textit{loadings}. Let $S \in \RR^{d \times d}$ denote the empirical covariance matrix and let $X \in \RR^{n \times d}$ denote our observed data matrix. $d$ is the number of variables while $n$ is the number of observations. The deconstruction procedure then goes as follows:
%\begin{enumerate}
%\item The vector of loadings corresponding to the first principal component, $\eta_1 \in \RR^d$, is chosen such that the component explains as much variance as possible. More precisely, for any unit vector $u \in \RR^d$ we find that $\V(u^T X^T)$ is maximized when $u = \eta_1$. Let $\lambda_1 = \V(\eta_1^T X^T)$.
%\item The second vector of loadings is then chosen as $\eta_2 = \argmax_{u: \, u \perp \eta_1} \V(u^T X^T)$ and we again define $\lambda_2 := \V(\eta_2 X)$.
%\item The third vector of loadings is $\eta_3 = \argmax_{u: \, u \perp \{\eta_1, \eta_2\}} \V(u^T X^T)$ and $\lambda_3 :=  \V(\eta_3^T X^T)$ \\
%\vdots
%\item[\refstepcounter{enumi} $d$.] The $d$th vector of loadings is $\eta_d = \argmax_{u: \, u \perp \{\eta_1, ..., \eta_{d-1}\}} \V(u^T X^T)$ and $\lambda_d = V(\eta_d^T X^T)$.
%\end{enumerate}
%The $\eta_i$s and $\lambda_i$s are determined uniquely (up to scaling) and they are given as the eigenvectors and eigenvalues of $S$, respectively. Moreover, we find that $X = \sum_{i=1}^d \eta_j \eta_j^T X$, so the procedure is really just a clever choice of rotation of the data, such that it is split up into vectors that are ordered by their variances, or more precisely, their variance contributions. We define the contribution to the total variance of the $k$th principal component as $\frac{\lambda_k}{tr(S)}$ and please note that, due to the properties of eigenvalue decomposition, $tr(S) = \sum_{i=1}^d \lambda_i$, which is the sum of the variances of the variables in the dataset.
%
% In this introduction, we have assumed full rank of the covariance matrix $S$ for simplicity, but it should be noted that this assumption is in no way needed for the results to hold.

\subsection{PCA-based data structure comparisons}
Above, we promised a method for intuitive, visual inspection of data structure similarities, but as of now, all intuition might have been lost in technicalities. The main point we want to emphasize from PCA is that whereas the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both data situations. This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variables are used differently in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

%Specifically, we can interpret the loadings as weights that determine the relative influence of each variable in each principal component.

%the two data sets will agree on which variables explain the most variance and also how this variance is explained.  Looking at the PCA loadings and the cumulative explained variance thus provides a straightforward non-parametric graphical approach for assessing similarity between two datasets.

In this paper we propose three diagnostic plots for comparing the loading patterns in two datasets $x_{11},\dotsc,x_{1 n_1} \in \RR^d$ and $x_{21},\dotsc,x_{2 n_2} \in \RR^d$ in the same $d$ variables. The construction of these plots, which we call \textit{CE plot}, \textit{angle plot} and \textit{chroma plot}, proceeds in the following steps:
\begin{enumerate}
\item Standardized the observations $x_{i1},\dotsc,x_{i n_i}$ to have zero mean and unit standard deviation separately for each of the $d$ variables within each datasets. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized observations, and form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.} 
\end{equation*}
\item Combine the standardized datasets into a joint dataset $\tilde{x}_1,\dotsc,\tilde{x}_n \in \RR^d$ with $n=n_1+n_2$ observations, standardize this dataset, and form the principal component analysis
\begin{equation*}
S = \frac{1}{n-2} \sum_{j=1}^n \tilde{x}_j \tilde{x}_j^\top = \sum_{k=1}^d \lambda_k \eta_k \eta_k^\top.
\end{equation*}
\item Make the diagnostic plot of interest. Details of the plots will be described below.
\item To investigate the sample variation randomly reallocate the combined and standardized dataset formed in step 2 to two datasets of $n_1$ and $n_2$ observations, respectively, and redo step 1 and 3. This is done several times, e.g.\ 10000 times, and compared to the plot for the original allocation.
\end{enumerate}

Before describing the details of the  \emph{CE plot}, the \emph{angle plot} and \emph{chroma plot}, we make some general remarks. First, it should be noted that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. \hl{Do we still need more about this topic? Then we can add something like:
\begin{itemize}
\item Following the greedy interpretation of PCA, we know that the first eigenvector is chosen as
$$\eta_1 = \argmax_{u \in \RR^d} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2  = \text{argmax}_{u \in \RR^d} \tilde{\V}(u^T X^T) $$
\item Now, imagine that $\V(x_1) >> \V(x_j)$ for all $j \in 2, ..., d$. Note that $(1, 0, ..., 0) X^T = x_i$. Then, clearly a vector $u$ that only picks out the first variable from the data matrix $X$, namely $u^T = (1, 0, ..., 0)$ will allow us to only look at the variance of $x_1$, which is very large. This has nothing to do with the interplay between variables, as the phenomenon will disappear if we scale $x_1$
\item A more formal explanation could come from this kind of intuition, if needed.
\end{itemize}
}
The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1. This also holds for the reallocated datasets, and in particular we have $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$. The random reallocations in step 4 provide resamples under the null hypothesis that the two sample populations have the same structure in the $d$ variables, except that the two populations are allowed to have the own standard deviations for each of the variables. This exception is induced by the initial step 1, which performs a standardization within the two datasets before they are combined in step 2. These resamples can be used for a visualization of the sample variation, and if a test statistic can be defined in step 3, then a $p$-value for the null hypothesis may be computed. Such $p$-values are know as permutation tests for Goodness-of-Fit (\hl{reference?}).

%In order to describe these plots we consider two different datasets in the same $d$ variables and with $n_1$ and $n_2$ observations, respectively. For dataset $i=1,2$ let $S_i \in \RR^{d \times d}$ be the empirical correlation matrix \hl{(find a way to introduce the empirial correlation matrix instead)}, and let $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0 $ and $\eta_{i1},\dotsc,\eta_{id} \in \RR^{d}$ be the corresponding variation components and loadings. The correlation matrices correspond to the covariance matrices for the variables after these have been standardized to unit standard deviation separately within the two datasets. Similarly, let $S \in \RR^{d \times d}$ be the empirical correlation matrix for the combined dataset with $n_1+n_2$ observations, and let $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ and $\eta_1,\dotsc,\eta_d \in \RR^d$ be the corresponding variation components and loadings.

\hl{To do: Something about what to do if not all variables are numerical. Anne: Maybe leave this for the discussion?
%Using correlation matrices is not the same as assuming standadized variables; there is a difference in the permutations.
}

\medskip

\subsubsection{The cumulative eigenvalue plot}
 The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the covariance matrix. 
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation. 
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{12}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
\end{align*}
This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}). Note that due to the standardization the last point will be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis the less alike the cumulative variation components for the two datasets are. \hl{Is this an informative x-axis? Should we maybe consider scaling it?}

>>>>

Classical statistical methodology is aimed at analyzing data from designed experiments and historically, statistical analyses have been done by researchers who knew the design and origin story of the data set well. However, the origin stories of data sets have changed over time and today a lot of data is accumulated without a specific purpose in mind. This is due to vast amounts of data being registered online and to a trend towards more open source research. The latter phenomenon in particular poses new challenges wrt.\ data quality assessment. When data are collected and made public without a specific end-point in mind, how do we ensure that differences in, say, choice measurement instruments, mode of administration, or sampling frame do not cause the data to be effectively divided into subsets that are simply not comparable? Three examples of this are

\begin{itemize}
\item Surveys often use mixed modes of administration, e.g. mail and telephone, and while this can improve response rates, the mode of administration can affect results \cite{Brambilla1987,McHorney1994} and differences in response behavior can lead to biased results. Powers, Mishra and Young \citeyearpar{Powers2005} report effects of mode of administration on changes in mental health scores that are of a magnitude that is considered to be clinically meaningful.
\item The rapid growth of web surveys, due to low cost, timeliness, and other factors, generate large data sources that lack a sampling frame of the general population. However, it can be problematic to combine online panels (pre-recruited profiled pools of respondents) with intercept samples (a pool of respondents obtained through banners, ads, or promotions) \cite{Liu2016}.
\item Large scale open source data sets such as the PISA data and ESS (European Social Survey) data. In these data sets, data analysts are far away from the data producers and problem-specific recommendations about potential instrument-induced challenges in the data sets are not available for the data analysts.
\end{itemize}

\hl{Maybe something here about what PCA is usually used for, e.g. dimension reduction, exploratory data analysis, ...}

Assume that we have two data sets with the same variables, but different observations represented as a data set with
a subset-inducing variable. Assume next that we wish to compare structures without specifying a model, or even a variable of interest. The central question is whether the two data sets can readily be combined for the purpose of later data analysis, or if the subset-inducing variable implies heterogeneity across the subset division.

Sophisticated methods for addressing this question are available when we are willing to assume a statistical model, but when these models are taken away, a remarkable void of methods is left behind. What is needed is a procedure that compares differences in overall data structures in two (or more) subsets of a data set without assuming neither directional nor hierarchical relationships between the variables. Using parametric models is not a general data structure comparison method, but rather a fitted-model comparison method. It addresses the interplay between the model and the data, not the data alone. 
Simple methods like variable-by-variable tests in distributional differences suffer from the drawback that they only address marginal differences and not to the interplay between variables. Entry-by-entry comparison of the two empirical correlation matrices quickly becomes unmanageable as the number of variables increase. For these methods significant $p$-values are likely to abound as the sample size increases. Parametric models using, e.g., latent variable models moves beyond the marginal approach, but need a pre-specified model.

We propose a suite of three new tools for this task, which we will refer to collectively as Principal Component Analysis-based Data Structure Comparisons (PCADSC). These methods employ the principal component decomposition of the empirical covariance matrix performed on two subsets of a dataset in order to create intuitive visualizations of data structure differences. Software for using these tools is available in an \texttt{R}-package. \hl{how to refer to our package?} This yields a solution that is largely independent of the sizes of the two subsets of data. Thereby, a lot of methods that compare each subset to the full data set in some sense are excluded.

This manuscript is structured as follows: First, in Section Section \ref{sec:pcadscintro}, we describe the PCADSC procedures, including a brief introduction to principal component analysis (PCA) in general. In Section \ref{sec:dataexample}, we present a worked data example using open source, online available data on psychological well-being in three European countries. More specifically, we compare data from Denmark, which has repeatedly been rewarded "happiest country in the world", with data from Bulgaria and Sweden, respectively, to investigate whether or not psychological well-being - and thus happiness - is really a concept that is universal beyond country borders, or if such rankings of happiness are inherently meaningless. At last, we discuss limitations and merits of the PCADSC tools in Section \ref{sec.Discussion}.

@@

A naive approach for data structure comparisons might therefore be to compute the empirical covariance matrices on each of the two data subsets and simply compare these matrices. Though the idea perhaps sounds appealing at first, it is quite difficult to assess similarity of matrices, and moreover, it becomes increasingly difficult for large numbers of variables and thus high dimensional covariance matrices. There is simply too much information to consider at once. However, by use of linear algebra, we can decompose and recompose the covariance matrix such that the distinct dimensions of information withheld in it are clearly separated and ordered according to their relative importance. In particular, we find a new representation of the covariance matrix that makes it possible to gain an overview of the most interesting aspects of the data. On such decomposition strategy is using \textit{principal component analysis} (PCA), which uses eigenvalue decomposition in order to obtain a new representation of the covariance matrix.

\subsection{A brief introduction to Principal component analysis}
Consider $n$ observations $x_1,\dotsc,x_n \in \RR^d$ of $d$ variables, let $\bar{x} = (\bar{x}_1, ..., \bar{x}_d)^\top = \frac{1}{n} \sum_{i=1}^n x_n$ denote their averages, and let $S = \frac{1}{n-1} \sum_{i=1}^n (x_i-\bar{x}) (x_i-\bar{x})^\top \in \RR^{d \times d}$ denote the empirical covariance matrix of the full data matrix, $X$. We assume in the following that all variables of $X$ have a numerical interpretation, e.g. by being continuous or ordinal and categorical. For a given $q \leq d$, principal component analysis (PCA) is a tool for finding a new representation of the dataset of dimension $q$ such that the least possible amount of information is lost. More specifically, we wish to minimize the loss when looking at a projection of $X$ onto a $q$-dimensional space, rather than the orignial $d$-dimensional one. Formally, we define the \emph{rank-q-reconstruction error} as the minimal squared error that is achievable by linear subspaces $K_q \subset \RR^d$ of dimension $q < d$, that is
\begin{equation*}
\min_{K_q} \sum_{i=1}^n \min_{z \in K_q} \lVert x_i - \bar{x} - z \rVert^2 =
\min_{K_q} \sum_{i=1}^n \lVert x_i - \bar{x} - \text{proj}_{K_q}(x_i - \bar{x}) \rVert^2.
\end{equation*}
and the theory of \emph{Principal component analysis} (PCA) not only ensures the existence of a subspace $\hat{K}_q \subset \RR^d$ that attains this minimum, it also provides an explicit description of $\hat{K}_q$ and the rank-q-reconstruction error.  \hl{Reference here? Maybe to like original Pearson article?}

More specifically, the rank-q-reconstruction error is attained when we choose
\begin{equation*}
\hat{K}_q = \text{span}\{\eta_1,\dotsc,\eta_q\}
\end{equation*}
where $\eta_1, ..., \eta_q$ are the $q$ first eigenvectors of the empirical covariance matrix, $S$, as ordered by the size of their associated eigenvalues, $\lambda_1, ..., \lambda_d$. In the PCA framework, we refer to the eigenvectors as \textit{loadings}. The eigenvalues may be understood as \textit{variance components}, as the sum of the marginal empirical variances is preserved under eigenvalue decomposition, that is,
$$\text{trace}(S) = \sum_{j=1}^d \hat{V}(X_j) = \sum_{j=1}^d \hat{V}(\eta_j^\top X^\top) = \sum_{j=1}^d \lambda_j$$
where $X_j \in \RR^n$ denotes the $j$th variable of $X$ and $\hat{V}$ is the empirical variance function. This again emphasizes that we do not change the covariance structure of a dataset when performing PCA; we merely use linear algebra to make it easier to describe. And as eigenvalues are uniquely defined, and eigenvectors are uniquely defined up to a change of sign whenever the eigenvalues are different, the representation hereby obtained is a valid object of inference. If $n < d$ or if the covariance matrix does not have full rank, it is possible to obtain non-unique eigenvalues, e.g.\ $\lambda_i=\lambda_{i+1}=\dotsm=\lambda_0$, but even in this case, the associated eigenvectors $\eta_i,\eta_{i+1},\dotsc,\eta_j$ are uniquely defined up to a common rotation.

The $j$th loading can also be found iteratively as the unit vector $u \in \RR^d$ orthogonal to $\hat{K}_{j-1}$ that maximizes the variation of the associated scores:
\begin{align*}
\eta_j &= \argmax_{u \in \RR^d\colon u \perp \hat{K}_{j-1}} \sum_{i=1}^n \lVert u^\top (x_i - \bar{x}) \rVert^2, &
\lambda_j &= \frac{1}{n-1} \sum_{i=1}^n \lVert \eta_j^\top (x_i - \bar{x}) \rVert^2.
\end{align*}
where the initial subspace is defined as $\hat{K}_0 = \{0\}$. It is worth emphasizing that this greedy approach of successively adding the next direction $\eta_j$ explaining most of the remaining variation, also gives the sequence $\hat{K}_q = \hat{K}_{q-1} \oplus \text{span} \{\eta_q\}$ of subspaces minimizing the rank-q-reconstruction error. This strong interpretation of PCA, which is often overlooked in the literature, means that the sequence of loadings $\eta_j$ and their associated variation components $\lambda_j$ yield a simultaneous description of the structure of the data set for all approximating dimensions $q$. This implies that the loadings and the variation components can be used to investigate the structure of the data set without the need to decide on an approximating dimension, $q$, a priori.

\subsection{Using PCA for data structure comparisons}
All in all, PCA qualifies as an appealing first step in structural comparisons of two datasets containing the same variables, and especially the loadings and variance components are meaningful quantities to compare across such different datasets. Usually, when performing PCA with other purposes in mind, the main interest lies in the \textit{scores}, i.e. the projections $\eta_j^\top (x_i - \bar{x})$ of the observations onto the loadings. But where the scores describe the observations, the variation components and the accompanying loadings describe the usage of the variables. If two different datasets with the same variables, but different samples of observations, have similar loading patterns, then the variables appear to be measuring the same underlying quantities in both datasets. This can be the case while the two sets of scores could be arbitrarily different, which e.g.\ could happen if the two datasets were taken from two different populations of subjects. On the other hand, if the loading patterns are different in the two datasets, then this indicates that the variables are used differently in the two data situations, and hence it would be criticizable to use these variables for comparisons across the two datasets.

The tools presented in this paper are all based on comparing the PCA results across two different datasets that contain the same variables. We denote these two datasets by $X_1$ and $X_2$, respectively, with $X_1$ consisting of $n_1$ observations of $d$ variables $x_{11},\dotsc,x_{1 n_1} \in \RR^d$, and similarly, $X_2$ consisting of $n_2$ observations of the $d$ variables, $x_{21},\dotsc,x_{2 n_2} \in \RR^d$. For each of these two datasets, we complete the following steps:
\begin{enumerate}
\item Standardized each of the variables to have mean zero and unit standard deviation. Let $\tilde{x}_{ij} \in \RR^d$ for $i=1,2$ and $j=1,\dotsc,n_i$ be the standardized datasets. 
\item Form the principal component analysis
\begin{equation*}
S_i = \frac{1}{n_i-1} \sum_{j=1}^{n_i} \tilde{x}_{ij} \tilde{x}_{ij}^\top = \sum_{k=1}^d \lambda_{ik} \eta_{ik} \eta_{ik}^\top \quad \text{for $i=1,2$.}
\end{equation*}
thereby obtaining loadings $\eta_{ik}$ and variance components $\lambda_{ik}$ for $i \in \{1, 2\}$ and $k \in 1, ..., d$.
\end{enumerate}
The hereby obtains PCA decompositions of the covariance matrices can then be compared. We present three diagnostics plots that are designed to shine a light on different types and nuances of data structure differences. These three plots are

\begin{description}
\item[The CE plot:] The CE (cumulative eigenvalue) plot can be used to illustrate differences in variance components, that is, in the relative importance of the directions identified by the PCA. The CE plot is accompanied by two permutation-tests, testing the hypothesis of no difference in variance components.
\item[The angle plot:] The angle plot compares both loadings and variance components at once and it can be used to understand the information loss if the data structure of one dataset is superimposed on the other, thereby revealing which principal components (i.e. loading and variance component pairs) that are most similar and most different across the two datasets
\item[The chroma plot:] The chroma plot is primarily an illustration of the loading patterns and it targets the question of how the roles of the original variables are different between the two datasets, thus leading the data structure comparison question back to its original, empirical context.
\end{description}
For the deepest understanding of the data structure differences in two datasets, we suggest using all three steps in the above order.

But before diving deeper into the details of the \emph{CE plot}, the \emph{angle plot} and the \emph{chroma plot}, a general remark about standardization in PCA is in place. We would like to emphasize that PCA is sensitive to scaling, as the procedure deconstructs the covariance matrix in components according to the most explained variance. This implies that if a variable has a very large sample variance (possibly because of its scale), this variable will be always be deemed highly influential, no matter the structure of the data. Therefore, the variables should always be scaled prior to performing PCA. Note that the covariance matrix for the standardized variables is the same as the correlation matrix for the original variables, so this simply corresponds to performing data structural comparisons of the correlation matrices rather than the covariance matrices. The standardization makes the variables comparable on the same scale, i.e.\ units of standard deviation, and it implies that the diagonal elements of $S$, $S_1$, and $S_2$ all equals 1, and thus also that $\sum_{k=1}^d \lambda_k = \sum_{k=1}^d \lambda_{1k} = \sum_{k=1}^d \lambda_{2k} =  d$.

\subsubsection{The cumulative eigenvalue plot}
 The cumulative eigenvalue (CE) plot compares the variation components, i.e. the eigenvalues of the covariance matrix.
%If these are the same in the two sample populations, then the best estimate for the variation components are $\lambda_1 \ge \dotsm \ge \lambda_d \ge 0$ found in the combined dataset. And we would expect $\lambda_{i1} \ge \dotsm \ge \lambda_{id} \ge 0$ for $i=1,2$ to be alike excepts sample variation.
In order to investigate whether the same proportion of the total variation can be described by the same number of principal components in the two datasets, we plot a piecewise linear curve connecting the points
\begin{align*}
(0,0), &&
(\lambda_1,\lambda_{11}-\lambda_{12}), &&
(\lambda_1 + \lambda_2,\lambda_{11}+\lambda_{12}-\lambda_{21}-\lambda_{22}), &&
\ldots, &&
\bigg( \sum_{j=1}^d \lambda_j, \sum_{j=1}^d \lambda_{1j} - \sum_{j=1}^d \lambda_{2j} \bigg).
\end{align*}
This may be seen as a cumulative Bland-Altman plot for the variation components (\hl{reference to cumulative residuals and to Bland-Altman}). Note that due to the standardization, the last point will always be equal to $(d,0)$. Thus, this curve will begin and end at the x-axis. And the larger excursions it makes away from the x-axis, the less alike the cumulative variation components for the two datasets are. Moreover, a positive cumulative differences implies that dataset 1 holds more information in the first components than dataset 2 does.

In order to test whether these cumulative differences are statistical artifacts or if they represent something real, we have implemented both the \emph{Kolmogorov-Smirnov} and the \emph{Cram\'er-von Mises} test statistics, which are given by
\begin{align*}
\text{KS} &= \max_{k=1,\dotsc,d} \bigg\lvert \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg\rvert, &
\text{CvM} &= \sum_{k=1}^{d-1} \frac{\lambda_k + \lambda_{k+1}}{2} \bigg( \sum_{j=1}^k \lambda_{1j} - \sum_{j=1}^k \lambda_{2j} \bigg)^2.
\end{align*}
We conduct the tests as \textit{permutation tests}, that is, by randomly reallocating the combined and standardized datasets into two new datasets of $n_1$ and $n_2$ observations, respectively, and then redoing the CE plot steps and recalculating the test statistics. This should be done a large (e.g. 10000) number of times. Then, a $p$-value is obtained by computing the proportion of reallocated datasets that lead to even larger test statistics than the one we found for the original datasets.

The permutation test results are also used to visualize the uncertainty of the CE curve in the plots. In the CE plots shown in the following section, we plot the observed curve together with 20 of the resampled curves, as well as a shaded region visualizing pointwise 95 \% coverage intervals. If the observed curve is very different from the resampled curves or if it is substantially outside the shaded region, then this also indicates differences between the two datasets.
